---
title: "Getting Started with kickstartR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with kickstartR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# Introduction

The `kickstartR` package helps you quickly initialize new R analysis projects with a standardized, organized directory structure and boilerplate files. This vignette will walk you through the main functionality and show you how to get started with creating well-organized R projects.

## Why use kickstartR?

When starting a new data analysis project, you often face the same organizational challenges:

- Where should I put my raw data?
- How should I organize my R scripts?
- Where should outputs like plots and tables go?
- What should I include in my `.gitignore` file?
- How can I make my project reproducible?

`kickstartR` solves these problems by providing a standardized project structure that follows R community best practices.

## Installation

You can install kickstartR from GitHub:

```{r eval=FALSE}
# Install devtools if you haven't already
install.packages("devtools")

# Install kickstartR from GitHub
devtools::install_github("SidhuK/kickstartR")
```

```{r setup}
library(kickstartR)
```

## Basic Usage

The main function in kickstartR is `initialize_project()`. Let's create a simple project:

```{r eval=FALSE}
# Create a new project in the current directory
initialize_project("MyAnalysisProject")
```

This creates a directory structure like this:

```
MyAnalysisProject/
├── 01_data/
│   ├── 01_raw/           # Original, immutable data
│   ├── 02_processed/     # Cleaned, transformed data
│   └── 03_external/      # External data sources
├── 02_scripts/           # R scripts and analysis code
│   └── 00_main_script.R  # Main analysis script template
├── 03_output/            # All analysis outputs
│   ├── 01_figures/       # Generated plots and figures
│   ├── 02_tables/        # Generated tables and summaries
│   └── 03_reports_rendered/ # Rendered reports (HTML, PDF)
├── 04_models/            # Saved model objects
├── 05_notebooks/         # R Markdown, Jupyter notebooks
├── README.md             # Project documentation
├── .gitignore            # Git ignore rules
├── .here                 # For use with the 'here' package
└── MyAnalysisProject.Rproj # RStudio project file
```

## Project Structure Explained

### Data Organization

The `01_data/` folder is organized into three subfolders:

- **`01_raw/`**: Store your original, immutable data files here. Never modify these!
- **`02_processed/`**: Save cleaned, transformed, or merged datasets here
- **`03_external/`**: Put data from external APIs, databases, or other sources here

### Scripts and Code

The `02_scripts/` folder contains your analysis code. The package creates a starter script `00_main_script.R` with best practice templates.

### Outputs

The `03_output/` folder is organized by output type:

- **`01_figures/`**: Save plots, charts, and visualizations
- **`02_tables/`**: Export summary tables, model results as CSV/Excel
- **`03_reports_rendered/`**: Store rendered HTML, PDF reports from R Markdown

### Models and Notebooks

- **`04_models/`**: Save fitted model objects (`.rds`, `.pkl` files)
- **`05_notebooks/`**: R Markdown files for exploratory analysis and reporting

## Customization Options

You can customize the project structure using various parameters:

### Specify a Different Location

```{r eval=FALSE}
# Create project in a specific directory
initialize_project("MyProject", path = "~/Documents/R_Projects")
```

### Exclude Certain Folders

```{r eval=FALSE}
# Create a simpler project without models and notebooks folders
initialize_project("SimpleProject",
                   include_models = FALSE,
                   include_notebooks = FALSE)
```

### Skip RStudio Project File

```{r eval=FALSE}
# Don't create .Rproj file (if you're not using RStudio)
initialize_project("MyProject", create_rproj = FALSE)
```

### Overwrite Existing Projects

```{r eval=FALSE}
# Overwrite an existing directory (use with caution!)
initialize_project("ExistingProject", overwrite = TRUE)
```

## Recommended Workflow

Here's how to use kickstartR in your typical analysis workflow:

### 1. Create Your Project

```{r eval=FALSE}
initialize_project("CustomerAnalysis2024")
```

### 2. Open in RStudio

Open the `.Rproj` file to set up your RStudio environment with the correct working directory.

### 3. Add Your Data

Place your raw data files in `01_data/01_raw/`. For example:

- `customer_data.csv`
- `sales_data.xlsx`
- `survey_responses.json`

### 4. Start Analyzing

Edit `02_scripts/00_main_script.R` or create new scripts. The template includes helpful patterns:

```r
# Load libraries
library(tidyverse)
library(here)

# Set base directory
here::i_am("02_scripts/00_main_script.R")

# Load data
raw_data <- read.csv(here::here("01_data", "01_raw", "customer_data.csv"))

# Process data
processed_data <- raw_data %>%
  filter(!is.na(important_column)) %>%
  mutate(new_column = old_column * 2)

# Save processed data
write.csv(processed_data,
          here::here("01_data", "02_processed", "cleaned_customers.csv"),
          row.names = FALSE)

# Create visualizations
my_plot <- ggplot(processed_data, aes(x = x_var, y = y_var)) +
  geom_point() +
  theme_minimal()

# Save outputs
ggsave(here::here("03_output", "01_figures", "customer_analysis.png"),
       my_plot, width = 10, height = 6, dpi = 300)
```

### 5. Use the `here` Package

The package creates a `.here` file to work seamlessly with the [`here` package](https://here.r-lib.org/), which provides robust file path management:

```{r eval=FALSE}
library(here)

# Always works regardless of working directory
data <- read.csv(here::here("01_data", "01_raw", "mydata.csv"))
ggsave(here::here("03_output", "01_figures", "plot.png"), my_plot)
```

## Best Practices

The project structure created by kickstartR embodies several best practices:

### 1. Separate Raw and Processed Data

- Keep original data immutable in `01_data/01_raw/`
- Save all processed/cleaned data to `01_data/02_processed/`
- This makes your workflow reproducible and prevents data loss

### 2. Organize Outputs by Type

- Separate figures, tables, and reports for easy navigation
- Use descriptive filenames with dates: `customer_analysis_2024-01-15.png`

### 3. Use Relative Paths

- The `.here` file enables robust path management
- Scripts work on any computer without path modification

### 4. Version Control Ready

- Includes sensible `.gitignore` patterns for R projects
- Excludes large files (outputs, models) by default
- Tracks code and documentation, not generated files

### 5. Self-Documenting Structure

- Numbered folders indicate typical workflow order
- Clear folder names explain their purpose
- README template provides project overview

## Advanced Tips

### Working with Teams

When sharing projects created with kickstartR:

1. **Git Repository**: Initialize git and push to GitHub/GitLab
2. **Dependencies**: Document required packages in your README
3. **Data Sharing**: Consider using `pins`, `targets`, or cloud storage for large datasets
4. **Environment**: Use `renv` for package version management

### Integration with Other Tools

kickstartR works well with other R project tools:

- **`targets`**: Put your `_targets.R` file in the project root
- **`renv`**: Initialize `renv` after creating the project structure
- **`testthat`**: Add tests to a `tests/` folder
- **`pkgdown`**: If converting to a package, the structure translates well

## Troubleshooting

### Common Issues

**Error: Directory already exists**

```{r eval=FALSE}
# Use overwrite = TRUE, but be careful!
initialize_project("MyProject", overwrite = TRUE)
```

**Want different folder names?**

Currently, kickstartR uses a fixed naming convention. If you need different folder names, you can:

1. Create the project with kickstartR
2. Manually rename folders as needed
3. Update paths in your scripts accordingly

**RStudio doesn't recognize the project?**

Make sure you're opening the `.Rproj` file, not just the folder. The `.Rproj` file tells RStudio this is a project.

## What's Next?

After creating your project structure with kickstartR:

1. **Explore your data** in `05_notebooks/` with R Markdown
2. **Develop your analysis** in `02_scripts/`
3. **Document your process** by updating the README
4. **Share your work** by pushing to GitHub and sharing the repository

The standardized structure makes it easy for others (including future you!) to understand and reproduce your work.

## Getting Help

If you encounter any issues or have suggestions:

- Check the [function documentation](../reference/initialize_project.html)
- Look at the [GitHub repository](https://github.com/SidhuK/kickstartR)
- Open an issue for bugs or feature requests

Happy analyzing! 🚀
